# Christian Santos Medellin
# R notebook to perform deseq2 analysis on abundance of vOTUs from spruce fieldsite

```{r}
source("../General/general_functions.R")
library(DESeq2)
library(biobroom)
library(tidyverse)
```


Load data, reformat, and filter out vOTUs from count table that are not present in the tpm table. 
```{r}
map <- read.table("../Data/metadata.csv", sep = ",", header = T) %>% 
  mutate(YearF = as.factor(Year),
         Group = paste(Year, Depth, sep = "."))

otu.tpm <- read.table("../Data/190807_all_otu_table.txt", sep = "\t", header = T)
row.names(otu.tpm) <- otu.tpm[,1]
otu.tpm <- otu.tpm[,-1]
otu.tpm <- otu.tpm[,match(map$SampleID, colnames(otu.tpm))]
otu.tpm <- otu.tpm[rowSums(otu.tpm)>0,]

good.votus <- otu.tpm>0

otu.count <- read.table("../Data/200922_counts_covtab_clean.csv", sep = ",", header = T)
row.names(otu.count) <- otu.count[,2]
otu.count <- otu.count[,-c(1,2,3)]
otu.count <- otu.count[match(rownames(good.votus), rownames(otu.count)),match(colnames(good.votus), colnames(otu.count))]
otu.count <- otu.count * good.votus
```



#Generate the DESeq object
```{r}
deseq <- DESeqDataSetFromMatrix(countData = otu.count,
                                colData = map,
                                design = ~ Depth + Year)

## Given that every vOTU contains at least one zero, we cannot use the default DESeq algorithm to estimate size factors. An alternative is to calculate the geometric means (https://support.bioconductor.org/p/62246/#62250)

cts <- counts(deseq)
geoMeans <- apply(cts, 1, function(row) if (all(row == 0)) 0 else exp(mean(log(row[row != 0]))))
deseq <- estimateSizeFactors(deseq, geoMeans=geoMeans)

deseq <- DESeq(deseq)
```

Test which vOTUs are significantly impacterd by depth. I am treating depth as a categorical factor with 4 levels and, instead of running pairwise wald tests, I am using an LRT approach to detect any variation across depth levels. 
```{r}
depth.dds <- DESeq(deseq, test = "LRT", reduced = ~ Year)
depth.res <- tidy(results(depth.dds)) %>%
  dplyr::rename("OTU_ID" = "gene")

depth.sig <- filter(depth.res, p.adjusted < 0.05)

depth.sig
```


Perform the clustering
```{r}
#this calculates the zscores of each OTU_ID across samples and then calculates the man zvalue for each depth
otu.zs <- otu.tpm %>%
  rel_ab() %>%
  tidy_otu() %>%
  filter(OTU_ID %in% depth.sig$OTU_ID) %>%     ######## this is the line that you need to comment out if you want to do the clustering with the whole dataset
  group_by(OTU_ID) %>%
  mutate(zValue = (Count - mean(Count))/sd(Count)) %>%
  inner_join(map, by = "SampleID") %>%
  group_by(depth, OTU_ID) %>%
  summarise(MeanZS = mean(zValue)) 

# In order to do the clustering, you need to format it as a matrix
otu.zs.matrix <- otu.zs %>%
  spread(key = OTU_ID, value = MeanZS) %>%
  as.data.frame()
row.names(otu.zs.matrix) <- otu.zs.matrix$depth
otu.zs.matrix <- otu.zs.matrix[,-1]
otu.zs.matrix <- as.matrix(otu.zs.matrix)
group.dist <- dist(otu.zs.matrix)
otu.dist <- dist(t(otu.zs.matrix))

#This part does the clustering
otu.hc <- hclust(as.dist(otu.dist), method = "ward.D") ## you can also play with the method you are using to cluster
otu.ord <- otu.hc$labels[otu.hc$order]
otu.ord <- data.frame(OTU_ID = otu.ord, order = 1:length(otu.ord))
otu.cut <- cutree(otu.hc[c(1,2,4)],k = 3) ### k will define how may clusters you are getting from your data. I used 3 because the PCoA data indicates that there are three clear groups: 10cm, 40cm, and both 100 and 150m
otu.clusters <- data.frame(OTU_ID = names(otu.cut),
                           Cluster = otu.cut) %>% 
  inner_join(otu.ord, by = "OTU_ID")

# plot
otu.zs %>% 
  inner_join(otu.clusters, by = "OTU_ID") %>% 
  ggplot(aes(as.factor(depth), reorder(OTU_ID, order), fill = MeanZS)) +
  geom_tile() +
  scale_fill_distiller(name = "Mean Rel. Abund.\n(z-score)",palette = "Greys") +
  facet_grid(Cluster ~ ., scales = "free", space = "free")
```

Calculate the hypergeometric test
```{r}
aquatic.votus <-  read.table("../Data/200922_vOTU_water_VC.csv", sep = ",", header = T)$SampleID
aquatic.votus

#Define how many OTUs in your 2699 detected were aquatic or not aquatic
aquatic.uni <- sum(row.names(otu.tpm) %in% aquatic.votus)
non.aquatic.uni <- sum(!row.names(otu.tpm) %in% aquatic.votus)

# Define how many OTUs in your detected clusters are aquatic or not aquatic
aquatic.clusters <- otu.clusters %>% 
  mutate(Type = ifelse(OTU_ID %in% aquatic.votus, "Aquatic", "NonAquatic")) %>% 
  group_by(Cluster, Type) %>% 
  dplyr::count() %>% 
  spread(key = Type, value = n, fill = 0) 
  

### Calculate the hypergeometric test per cluster
aquatic.nest <- aquatic.clusters %>% 
  group_by(Cluster) %>% 
  nest 
  

## This is the function that runs the hypergeometric test. You need to use lower.tail=F to calculate overrepresentation rather than underrepresentation (https://seqqc.wordpress.com/2019/07/25/how-to-use-phyper-in-r/)
get_hype <- function(x){
  phyper(x$Aquatic, aquatic.uni, non.aquatic.uni, x$Aquatic + x$NonAquatic, lower.tail = F)
}

## hgt is the pvalue
aquatic.nest %>% 
  mutate(hgt = map(data, get_hype)) %>% 
  unnest(hgt)
```

Two ways to visualize how the proportion of aquatic vOTUs in each cluster compares to the proportion observed in the whole dataset
```{r}
aquatic.clusters %>% 
  ungroup() %>% 
  mutate(Cluster = as.character(Cluster)) %>% 
  add_row(Cluster = "Total", Aquatic = aquatic.uni, NonAquatic = non.aquatic.uni) %>% 
  gather(key = "Type", value = "nOTUs", -Cluster) %>% 
  ggplot(aes(Cluster, nOTUs, fill = Type)) +
  geom_bar(stat = "identity", position = "fill")

aquatic.clusters %>% 
  ungroup() %>% 
  mutate(Cluster = as.character(Cluster)) %>% 
  add_row(Cluster = "Total", Aquatic = aquatic.uni, NonAquatic = non.aquatic.uni) %>% 
  group_by(Cluster) %>% 
  mutate(PercentAquatic = Aquatic/sum(Aquatic + NonAquatic)) %>% 
  ggplot(aes(Cluster, PercentAquatic)) +
  geom_bar(stat = "identity")
```

